---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PMC

<!-- badges: start -->
<!-- badges: end -->

The PMC package provides tools to calculate the Peat Moisture Code (PMC)—a daily moisture index designed specifically for peatland ecosystems with deep organic soil profiles.
PMC integrates effective precipitation, evapotranspiration, and peat hydrological properties to simulate subsurface moisture conditions relevant for fire danger rating, drought monitoring, and ecological assessment.

The package includes:

*Functions to prepare and standardize meteorological inputs

*Tools to compute effective precipitation, PET, and daily PMC

*Visualization utilities

*Four example datasets from peatland-dominated regions in Canada

## Installation

You can install the development version of PMC from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ElizabethArangoRuda/PMC")
```

## Example

Load the package:

```{r c1}
library(PMC)
```

## Example datasets

The package includes four peatland meteorological datasets:

*Fort_McMurray.rda
*Grande_Prairie.rda
*Mildred_Lake.rda
*Slave_Lake.rda

```{r c2}
data("Mildred_Lake")
summary(Mildred_Lake)
```
## Preprocessing functions

clean_data()

Prepares input data for PMC calculations by standardizing column names and formats.
It is recommended to run this function before any other PMC-related functions.

```{r c3, echo = FALSE}
df <- clean_data(Mildred_Lake)
```
mm_to_cm()

Converts selected variables from millimeters (mm) to centimeters (cm) by multiplying by 0.1.

```{r c4, echo = FALSE}
?mm_to_cm
df <- mm_to_cm(df, columns = "PPT")
```
eff_ppt()

Calculates effective precipitation, representing the portion of precipitation that contributes to peat moisture storage (after subtracting negligible rainfall events).
The threshold argument defines the minimum precipitation to be considered effective.

```{r c5, echo = FALSE}
?eff_ppt
df <- eff_ppt(df, column = "PPT_cm", threshold = 0.1, year_to_plot = "2015")
```
PETpt()

Computes daily potential evapotranspiration (PET) using the Priestley–Taylor method, a simplified alternative to the Penman–Monteith equation suitable for peatland environments. 

```{r c6, echo = FALSE}
?PETpt
df <- PETpt(df,
  latitude = 65.2825,
    alpha = 1,
    y = 0.063,
    Gsc = 0.0820,
    lambda = 2453,
    a = 0.17,
    b = 0.59,
    year_to_plot = "2014")
```

## Computing the Peat Moisture Code (PMC)

The PMC() function calculates the daily Peat Moisture Code using a water‐balance approach that integrates effective precipitation and actual evapotranspiration.
PMC increases during drying periods and decreases when water storage is replenished.

```{r c7, echo = FALSE}
?PMC
df <- PMC(df,
    PET_column = "PET",
    A = 0.8674,
    B = 0.0540,
    start_PMC = 10,
    C = 0.15, # C = 0.1
    Sy_min = 0.1,
    PMC_min = 0.1,
    year_to_plot = "2014")
```
## Visualization utilities

The package includes visuals(), a helper function to produce a four-panel figure summarizing key hydrometeorological variables.

```{r c8, echo = FALSE}
?PMC
df <- visuals(df,
    var1 = "PET",
    var2 = "Tavg",
    var3 = "PMC",
    var4 = "eff_Precip_cm",
    x_label_vr1 = "PET (cm)",
    x_label_vr2 = "Tair (°C)",
    x_label_vr3 = "PMC",
    x_label_vr4 = "Effective Precipitation (cm)",
    year_to_plot = "2014"
  )

```
